Neutral models of \textit{de novo} gene emergence suggest that gene evolution has a preferred trajectory

### 2023-02-27 Manuscripts/Response_to_Reviewers_Ravi2023MBE.tex

We thank the reviewer for this important suggestion. We now calculate the probability of fixation as well as the probability of extinction under the assumptions of (1) total neutrality, (2) weak positive selection, and (3) weak negative selection. We include these analyses in the new figure 1, and describe the rationale (lines 207 -- 232) and the method (lines 796 -- 798, new table 2) in the main text.

We now explicitly state that fixation under total neutrality depends solely on the mutation rate (lines 221 -- 222). We have also mentioned that time to fixation would be much smaller than time to gain a gene (lines 230 -- 232)

{\itshape Another difficulty for me was the argument about whether the existence of a lineage specific gene really implies it evolved only in that lineage. The alternative being an earlier gain with losses in all lineages but one. The paper justifies this by deriving that the probability of losing a de novo gene is much larger than gaining it. I struggle to see how that comparison alone is sufficient to make this argument. It seems to me one would have to tally up all relevant probabilities of gaining, losing, and keeping the gene along the relevant internal and terminal branches of the species tree. The length of these branches also seems like it might matter a lot. A gain-with-2-losses trajectory, like the authors suggest may be likely, requires the new gene to be gained and then kept along one internal and one terminal branch. It would seem to me that this is vastly less likely under strict neutrality than a lineage specific gain. I may of course be wrong, but I think a very simplistic argument based on only two probabilities with no treatment of the actual phylogeny is not sufficient to convince me of the authors' argument.}

We are not sure if we understood the reviewer's point completely. We do agree that phylogenetic inferences should take into consideration branch lengths. We assume that branch lengths should have minimal effect when comparing gene gains and losses in populations that haven't diverged from each other over a long time. We now explicitly say that we analyse populations (lines 148 -- 154) and not species that could have diverged over many millions of years. Thus our model is suitable for relatively shorter timescales. Furthermore, multiple gene gain and loss events within a lineage or branch would be highly rare, given the low probability of these events.

@@

Thus given an observation that a population has a gene but two other populations do not have it, there can be two assumptions:

1. Gene is gained in one population
2. Gene was present in the ancestor and lost in two populations

The parsimonious assumption would be that gene is only gained in one population. What we show is that the gene \textbf{can} be lost in two populations independently. It is indeed possible that these two populations in turn share a common ancestor that had already lost the gene. This event is more probable than two gene losses. This in turn strengthens our argument that gene losses are much more frequent than gene gain. Multiple independent losses is an extreme case that may be less feasible than a single loss followed by population divergence. Nevertheless it is possible, and can be more likely than a single gene gain event. We hope we have clarified the reviewer's concern.

### 2023-02-27 Manuscripts/deNovoRates.tex
modified:   Manuscripts/deNovoRates.tex

\begin{equation}
P_\textit{RNA-gain}(k) = P_\textit{RNAinit} \times \begin{pmatrix}
P_\textit{polyA-gain}\times (P_\textit{nopolyA-stay})^{3k-5} \\[1em]
 +\  P_\textit{polyA-stay}\times (P_\textit{nopolyA-stay})^{3k-5} \times (3k-6) \times P_\textit{polyA-loss}
\end{pmatrix}  
\label{eqrnagain}
\end{equation}

@@

\begin{equation}
P_\textit{RNA-stay}(k) = P_\textit{polyA-stay} \times (P_\textit{nopolyA-stay})^{(3k-5)}
\label{eqrnastay}
\end{equation}
}
\subsection{Interdependence of transcription and ORF probabilities}

>>>>

\begin{equation}
P_\textit{RNA-gain}(k) = P_\textit{RNAinit} \times \begin{pmatrix}
P_\textit{polyA-gain}\times (P_\textit{nopolyA-stay})^{3k-5} \\[1em]
 +\  P_\textit{polyA-stay}\times (P_\textit{nopolyA-stay})^{3k-6} \times (3k-6) \times P_\textit{polyA-loss}
\end{pmatrix}  
\label{eqrnagain}
\end{equation}

@@

\begin{equation}
P_\textit{RNA-stay}(k) = P_\textit{RNAinit} \times P_\textit{polyA-stay} \times (P_\textit{nopolyA-stay})^{(3k-5)}
\label{eqrnastay}
\end{equation}
}
\subsection{Interdependence of transcription and ORF probabilities}
